Exceedance probability comparison for the simplified Naomi model

### 2023-06-20 src/docs_21-06-23-mlgh/21-06-23-mlgh.Rmd
Mostly working on MLGH presentation, some other writing

```{r, echo=FALSE, fig.cap="Naomi is a district-level model of HIV indicators.", out.width = '90%'}
knitr::include_graphics("depends/naomi_results.png")
```

@@

```{r, echo=FALSE, fig.cap="Web interface promotes data ownership, data use, and data quality. From \\texttt{https://naomi.unaids.org/.}", out.width = '70%'}
knitr::include_graphics("naomi_user.png")
```

# Better estimates by triangulating information

@@

# The model has a big Gaussian latent field

@@

If you input $\theta_1, \ldots, \theta_{24}$ then it'll return a 467 length mean vector $\x(\btheta)$ and $467 \times 467$ length covariance matrix $\hat{\mathbf{H}} (\btheta)$

@@

@@

1. Further directions for this research $\color{lolit}{\text{including my suggestions for short masters or PhD projects}}$  
2. Impactful academic or industry jobs using Bayesian statistics $\color{lolit}{\text{to begin around the end of this year when I graduate (I hope!)}}$

@@

* Code and notebooks: `github.com/athowes/elgm-inf`
* Working paper on the way^[For sufficiently vague definition of "on the way"]: Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model $\color{lolit}{\text{Adam Howes, Alex Stringer, Seth Flaxman, Jeff Eaton}}$

@@

<!-- # Two deterministic methods -->

<!-- * We use two deterministic^[In contrast to the most famous approximate Bayesian inference method, Markov chain Monte Carlo, which is fundamentally stochastic.] methods to approximate our integral -->

<!-- 1. The Laplace approximation -->
<!-- 2. Quadrature -->

@@

<!-- # Which parameters should we treat as Gaussian? -->

<!-- * In spatio-temporal statistics we have data indexed by space and time -->
<!-- * We use random effects also indexed by space and time to model this data -->
<!-- * Spatio-temporal fields can be big -->
<!-- * Willing to make assumptions about how things vary over spacetime -->

<!-- $\implies$ treat the latent field parameters as Gaussian! That's the majority of the integral done. -->

@@

<!-- # Quadrature -->

<!-- * This brings us to our other method for solving integrals deterministically -->
<!-- * Say we have a function, then quadrature has two ingredients -->

<!-- 1. Nodes: points to evaluate the function at -->
<!-- 2. Weights: importance of function evaluation at that point -->

<!-- * You evaluate the function at the nodes, then do a weighted sum to calculate your integral -->

<!-- # Trapezoid rule example -->

<!-- * Let's compute $\int_0^\pi \sin(x) \text{d}x = 2$ using quadrature -->

<!-- ```{r} -->
<!-- trapezoid_rule <- function(x, spacing) { -->
<!--   # Assumes nodes are evenly spaced -->
<!--   w <- rep(spacing, length(x)) # Weights given by space between nodes -->
<!--   w[1] <- w[1] / 2 # Apart from the first which is halved -->
<!--   w[length(x)] <- w[length(x)] / 2 # And the last, also halved -->
<!--   sum(w * x) # Compute the weighted sum -->
<!-- } -->
<!-- ``` -->

<!-- ```{r message=FALSE, echo=FALSE,  fig.cap="With 10 nodes it's 0.02 off."} -->
<!-- finegrid <- seq(0, pi, length.out = 1000) -->

<!-- plot <- data.frame(x = finegrid, y = sin(finegrid)) %>% -->
<!--   ggplot(aes(x = x, y = y)) + -->
<!--    geom_line(col = cbpalette[1]) + -->
<!--    theme_minimal() + -->
<!--    labs(x = "x", y = "sin(x)") -->

<!-- trapezoid_plot <- function(N) { -->
<!--   grid <- seq(0, pi, length.out = N) -->
<!--   int <- trapezoid_rule(x = sin(grid), spacing = grid[2] - grid[1]) -->

<!--   plot +  -->
<!--     geom_bar( -->
<!--       data = data.frame(x = grid, y = sin(grid)), -->
<!--       aes(x = x, y = y), alpha = 0.7, stat = "identity", -->
<!--       inherit.aes = FALSE, fill = cbpalette[2]) + -->
<!--     theme_minimal() + -->
<!--     labs( -->
<!--       subtitle = paste0("Number of nodes: ", N, "\nTrapezoid rule estimate: ", round(int, 3), "\nTruth: 2"), -->
<!--       x = "x", y = "sin(x)" -->
<!--     ) -->
<!-- } -->

<!-- trapezoid_plot(N = 10) -->
<!-- ``` -->

<!-- ```{r message=FALSE, echo=FALSE, fig.cap="With 30 nodes it's 0.002 off."} -->
<!-- trapezoid_plot(N = 30) -->
<!-- ``` -->

<!-- ```{r message=FALSE, echo=FALSE, fig.cap="With 100 nodes it's pretty much correct."} -->
<!-- trapezoid_plot(N = 100) -->
<!-- ``` -->

<!-- # Adaptive Gauss-Hermite quadrature -->

<!-- * \textcolor{hilit}{Gauss-Hermite quadrature} is a method for picking nodes and weights based on the theory of polynomial interpolation -->
<!-- * It works especially well for statistical problems where the integrand looks like something multiplied by a Gaussian distribution -->
<!-- * The \textcolor{hilit}{adaptive} part means the nodes and weights are changed depending on the integrand -- this makes sense, especially when the integrand is also a function of the data -->
<!-- * Implemented by the `aghq` package [@stringer2021implementing] -->

<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Unadapted points in two dimensions with $k = 3$."} -->
<!-- mu <- c(1, 1.5) -->
<!-- cov <- matrix(c(2, 1, 1, 1), ncol = 2) -->

<!-- obj <- function(theta) { -->
<!--   mvtnorm::dmvnorm(theta, mean = mu, sigma = cov) -->
<!-- } -->

<!-- grid <- expand.grid( -->
<!--   theta1 = seq(-2, 5, length.out = 700), -->
<!--   theta2 = seq(-2, 5, length.out = 700) -->
<!-- ) -->

<!-- ground_truth <- cbind(grid, pdf = obj(grid)) -->

<!-- plot0 <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) + -->
<!--   geom_contour(col = cbpalette[1]) + -->
<!--   coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) + -->
<!--   labs(x = "", y = "") + -->
<!--   theme_minimal() -->

<!-- gg <- mvQuad::createNIGrid(2, "GHe", 3) -->

<!-- add_points <- function(plot0, gg) { -->
<!--   plot0 + -->
<!--     geom_point( -->
<!--       data = mvQuad::getNodes(gg) %>% -->
<!--               as.data.frame() %>% -->
<!--               mutate(weights = mvQuad::getWeights(gg)),  -->
<!--       aes(x = V1, y = V2, size = weights), -->
<!--       alpha = 0.8, -->
<!--       col = cbpalette[2], -->
<!--       inherit.aes = FALSE -->
<!--     ) + -->
<!--     scale_size_continuous(range = c(1, 2)) -->
<!-- } -->

<!-- add_points(plot0, gg) + -->
<!--   labs(size = "Weight", caption = "") -->
<!-- ``` -->

<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Add the mean $z + \\hat \\theta$."} -->
<!-- gg2 <- gg -->
<!-- mvQuad::rescale(gg2, m = mu, C = diag(c(1, 1)), dec.type = 2) -->

<!-- add_points(plot0, gg2) + -->
<!--   labs(size = "Weight", caption = "") -->
<!-- ``` -->

<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="First option: rotate by the lower Cholesky $Lz + \\hat \\theta$."} -->
<!-- gg3 <- gg -->
<!-- mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2) -->

<!-- add_points(plot0, gg3) + -->
<!--   labs(size = "Weight", caption = "") -->
<!-- ``` -->

<!-- ```{r, echo=FALSE, message=FALSE, fig.cap="Second option: rotate using the eigendecomposition $E \\Lambda^{1/2} z + \\hat \\theta$."} -->
<!-- gg3 <- gg -->
<!-- mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 1) -->

<!-- add_points(plot0, gg3) + -->
<!--   labs(size = "Weight", caption = "") -->
<!-- ``` -->

>>>>

```{r, echo=FALSE, fig.cap="Naomi is a district-level model of HIV indicators, like prevalence, treatment coverage, and incidence.", out.width = '90%'}
knitr::include_graphics("depends/naomi_results.png")
```

@@

```{r, echo=FALSE, fig.cap="Estimate generation via a web interface promotes data ownership, data use, and data quality. From \\texttt{https://naomi.unaids.org/.}", out.width = '70%'}
knitr::include_graphics("naomi_user.png")
```

# Better estimates by integration information from many sources

@@

# The model has a big, structured, Gaussian latent field $\x$

* Fixed effects $\color{lolit}{\text{Gaussian}}$
* Age structure $\color{lolit}{\text{AR1}}$
* Spatial structure $\color{lolit}{\text{IID, ICAR, BYM2}}$

Concatenate together as $\x$, length 467

# Smaller, non-Gaussian, hyperparameters $\btheta$

* Standard deviations $\color{lolit}{\text{Half-Gaussian}}$
* BYM2 proportion parameters $\color{lolit}{\text{Beta}}$
* AR1 autocorrelation parameters $\color{lolit}{\text{Uniform}}$

Concatenate together as $\btheta$, length 24

@@

If you input 24 length $\btheta$ then it'll return a 467 length mean vector $\hat \x(\btheta)$ and $467 \times 467$ length covariance matrix $\hat{\mathbf{H}} (\btheta)$.
Mean calculated using gradient based optimisation.
Gradients, and Hessian, obtained using automatic differentiation.

@@

# Use adaptive Gauss-Hermite quadrature to integrate over $\btheta$

Quadrature method based on the theory of polynomial interpolation which:

1. Works well when the integrand looks like a polynomial times a Gaussian
2. Adapts to the particular integrand based on the mode and Hessian
3. Is implemented by the `aghq` package [@stringer2021implementing]

```{r, echo=FALSE, message=FALSE, fig.cap="Unadapted Gauss-Hermite nodes in two dimensions with $k = 3$."}
figA1
```

```{r, echo=FALSE, message=FALSE, fig.cap="Add the mode $\\mathbf{z} + \\hat \\btheta$."}
figA2
```

```{r, echo=FALSE, message=FALSE, fig.cap="You could rotate by the lower Cholesky $\\mathbf{Lz} + \\hat \\btheta$."}
figA3
```

```{r, echo=FALSE, message=FALSE, fig.cap="Or you could rotate using the eigendecomposition $\\mathbf{E} \\bm{\\Lambda}^{1/2} \\mathbf{z} + \\hat \\btheta$."}
figA4
```

# 24 hyperparameters is too many for a dense grid

$k = 3$ points in 24 dimensions is not feasible
$$
3 \times 3 \times \cdots \times 3 = 3^{24}
$$
So we need to find something smaller $\color{lolit}{\text{(that still does a good job!)}}$

```{r, echo=FALSE, message=FALSE, fig.cap="An obvious thing to try is only keeping points from the largest eigenvectors: we call this PCA-AGHQ. Corresponds to variable choice of $k$ by dimension."}
figA5
```

```{r, echo=FALSE, message=FALSE, fig.cap=".", out.width = '90%'}
knitr::include_graphics("depends/tv-plot.png")
```

```{r, echo=FALSE, message=FALSE, fig.cap=".", out.width = '90%'}
knitr::include_graphics("depends/reduced-rank-plot.png")
```

# Yes, but did it work?

Run NUTS^[For 3 days!] as gold-standard, then compare to TMB to PCA-AGHQ using:

1. Marginal distributions $\color{lolit}{\text{point estimates, ECDF e.g. KS or AD}}$  
2. Joint distributions $\color{lolit}{\text{PSIS, MMD}}$ 
3. Policy relevant outcomes $\color{lolit}{\text{second 90, high incidence}}$

```{r, echo=FALSE, message=FALSE, fig.cap="PCA-AGHQ modestly improves estimation of the posterior mean.", out.width = '80%'}
knitr::include_graphics("depends/mean.png")
```

```{r, echo=FALSE, message=FALSE, fig.cap="PCA-AGHQ substantially improves estimation of the posterior standard deviation. TMB systematically underestimates, which you'd expect.", out.width = '80%'}
knitr::include_graphics("depends/sd.png")
```

```{r, echo=FALSE, message=FALSE, fig.cap="Strata probabilities of having greater than 81\\% ART coverage, and as such meeting the second 90 target. Both approximate methods are inaccurate for females.", out.width = '85%'}
knitr::include_graphics("depends/exceedance-second90.png")
```

```{r, echo=FALSE, message=FALSE, fig.cap="Strata probabilities of having greater than 1\\% HIV incidence, and as such being classified high incidence. Again, both approximate methods are inaccurate.", out.width = '85%'}
knitr::include_graphics("depends/exceedance-1inc.png")
```

# Further improvements look possible

1. Fix issues with scaling $\color{lolit}{\text{Logit scaled not uniformly more important than log scaled}}$  
2. Take into account importance for outputs of interest $\color{lolit}{\text{Variance of inputs isn't really what we care about}}$  
3. Take into account marginal skewness $\color{lolit}{\text{The more skewed, the more we should be placing lots of points}}$

```{r, echo=FALSE, message=FALSE, fig.cap="On the real scale, $[0, 1]$ hyperparameters appear to have more marginal variance than $[0, \\infty)$ hyperparameters. This doesn't really make them more important though.", out.width = '80%'}
knitr::include_graphics("depends/marginal-sd.png")
```

@@

1. Further directions for this research $\color{lolit}{\text{e.g. suggestions for short masters or PhD projects}}$  
2. Impactful academic or industry jobs using Bayesian statistics $\color{lolit}{\text{to begin around the end of this year when I graduate (hopefully!)}}$

| Method      | Details          |
| ----------- | ---------------- |
| Zulip       | Adam Howes       |
| Email       | `ath19@ic.ac.uk` |
| Calendly    | `adamthowes`     |

@@

* Code and notebooks: `github.com/athowes/elgm-inf`
* Working paper on the way^[For sufficiently vague definition of "on the way"], Any early readers greatly appreciated!
  * Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model $\color{lolit}{\text{Adam Howes, Alex Stringer, Seth Flaxman, Jeff Eaton}}$

@@

@@

@@

### 2023-06-20 src/docs_paper/paper.Rmd
Mostly working on MLGH presentation, some other writing

```{r results='hide', echo = FALSE}
cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")

@@

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 320,
  cache = TRUE,
  out.width = "95%",
  fig.align = 'center'
)

source("figures.R")
```

@@

```{r mean-sd, fig.cap="TMT."}
knitr::include_graphics("depends/mean-sd-comparison.png")
```

>>>>

```{r echo = FALSE}
cbpalette <- c("#56B4E9","#009E73", "#E69F00", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")

@@

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dpi = 320,
  cache = TRUE,
  out.width = "95%",
  fig.align = 'center'
)
```

```{r results='hide'}
source("figures.R")
```

@@

```{r mean-sd, fig.cap="TMT."}
knitr::include_graphics("depends/mean-sd.png")
```

### 2023-06-20 src/naomi-simple_exceedance/exceedance.Rmd
Mostly working on MLGH presentation, some other writing

ggsave("exceedance.png", exceedance_manuscript_plot, h = 6.5, w = 6.25, bg = "white")
```

>>>>

ggsave("exceedance.png", exceedance_manuscript_plot, h = 6.5, w = 6.25)

#' Separate versions are nice for presentations etc.

exceedance_manuscript_plot_second90 <- df_wide %>%
  filter(indicator == "Second 90") %>%
  pivot_longer(cols = c("TMB", "aghq"), names_to = "method", values_to = "estimate") %>%
  mutate(
    method = fct_reorg(method, "TMB" = "TMB", "PCA-AGHQ" = "aghq")
  ) %>%
  ggplot(aes(x = tmbstan, y = estimate)) +
    geom_point(aes(col = sex), shape = 1, alpha = 0.6) +
    coord_fixed(ratio = 1) +
    scale_colour_manual(values = c("#56B4E9", "#999999"), labels = c("Female", "Male")) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    geom_text(
    data = df_metrics %>%
      filter(indicator == "Second 90") %>%
      mutate(method = fct_relevel(method, "TMB", "PCA-AGHQ")),
      aes(x = -Inf, y = Inf, label = label),
      size = 3, hjust = 0, vjust = 1.5
    ) +
    facet_grid(indicator ~ method) +
    labs(x = "NUTS", y = "", col = "Sex") +
    theme_minimal()

ggsave("exceedance-second90.png", exceedance_manuscript_plot_second90, h = 4, w = 6.25)

exceedance_manuscript_plot_1inc <- df_wide %>%
  filter(indicator == "High incidence") %>%
  pivot_longer(cols = c("TMB", "aghq"), names_to = "method", values_to = "estimate") %>%
  mutate(
    method = fct_reorg(method, "TMB" = "TMB", "PCA-AGHQ" = "aghq")
  ) %>%
  ggplot(aes(x = tmbstan, y = estimate)) +
    geom_point(aes(col = sex), shape = 1, alpha = 0.6) +
    coord_fixed(ratio = 1) +
    scale_colour_manual(values = c("#56B4E9", "#999999"), labels = c("Female", "Male")) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
    geom_text(
    data = df_metrics %>%
      filter(indicator == "High incidence") %>%
      mutate(method = fct_relevel(method, "TMB", "PCA-AGHQ")),
      aes(x = -Inf, y = Inf, label = label),
      size = 3, hjust = 0, vjust = 1.5
    ) +
    facet_grid(indicator ~ method) +
    labs(x = "NUTS", y = "", col = "Sex") +
    theme_minimal()

ggsave("exceedance-1inc.png", exceedance_manuscript_plot_1inc, h = 4, w = 6.25)
```
